Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-17 ms.tex
fixed alexei's name
Jeff Oishi's changes.

\title{Enzo: An Adaptive Mesh Refinement Code for Astrophysics}
\author{The Enzo Collaboration:}
\author{Greg L. Bryan\altaffilmark{1}}
\author{Michael L. Norman\altaffilmark{2,3}}
\author{Brian W. O'Shea\altaffilmark{4,5}}
\author{Tom Abel\altaffilmark{6, 19}}
\author{John H. Wise\altaffilmark{7}}
\author{Matthew J. Turk\altaffilmark{1}}
\author{Daniel R. Reynolds\altaffilmark{8}}
\author{David C. Collins\altaffilmark{9}}
\author{Peng Wang\altaffilmark{6}}
\author{Samuel W. Skillman\altaffilmark{10,11}}
\author{Britton Smith\altaffilmark{4}}
\author{Robert P. Harkness\altaffilmark{12}}
\author{James Bordner\altaffilmark{2}}
\author{Ji-hoon Kim\altaffilmark{13}}
\author{Michael Kuhlen\altaffilmark{14,15}}
\author{Hao Xu\altaffilmark{2}}
\author{Nathan Goldbaum\altaffilmark{15}}
\author{Cameron Hummels\altaffilmark{16}}
\author{Alexei Kritsuk\altaffilmark{2}}
\author{Elizabeth Tasker\altaffilmark{17}}
\author{Stephen Skory\altaffilmark{2}}
\author{Christine Simpson\altaffilmark{1}}
\author{Oliver Hahn\altaffilmark{18}}
\author{Jeffrey S. Oishi\altaffilmark{6}}
\author{Geoffrey C So\altaffilmark{2}}
\author{Fen Zhao\altaffilmark{19}}
\author{Renyue Cen\altaffilmark{20}}
\author{Yuan Li\altaffilmark{1}}

\altaffiltext{1}{Columbia University, Department of Astronomy, New York, NY, 10025, USA}
\altaffiltext{2}{CASS, University of California, San Diego, 9500 Gilman Drive La Jolla, CA 92093-0424, USA}
\altaffiltext{3}{SDSC, University of California, San Diego, 9500 Gilman Drive La Jolla, CA 92093-0505, USA}
\altaffiltext{4}{Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\altaffiltext{5}{Lyman Briggs College and Institute for Cyber-Enabled Research, Michigan State University, East Lansing, MI 48824, USA}
\altaffiltext{6}{Kavli Institute for Particle Astrophysics and Cosmology, Stanford University, Menlo Park, CA 94025, USA}
\altaffiltext{7}{Center for Relativistic Astrophysics, School of Physics, Georgia Institute of Technology, 837 State Street, Atlanta, GA, USA}
\altaffiltext{8}{Southern Methodist University, 6425 Boaz Ln, Dallas, TX 75205, USA}
\altaffiltext{9}{Theoretical Division, Los Alamos National Laboratory, Los Alamos, NM 87544}
\altaffiltext{10}{Center for Astrophysics and Space Astronomy, Department of Astrophysical and Planetary Science, University of Colorado, Boulder, CO 80309, USA}
\altaffiltext{11}{DOE Computational Science Graduate Fellow}
\altaffiltext{12}{NICS, Oak Ridge National Laboratory, 1 Bethel Valley Rd, Oak Ridge, TN 37831, USA}
\altaffiltext{13}{Department of Astronomy and Astrophysics, University of California, Santa Cruz, CA 95064, USA}
\altaffiltext{14}{Theoretical Astrophysics Center, University of California Berkeley, Hearst Field Annex, Berkeley, CA 94720, USA}
\altaffiltext{15}{Department of Astronomy and Astrophysics, University of California, 1156 High Street, Santa Cruz, CA 95064, USA}
\altaffiltext{16}{Department of Astronomy/Steward Observatory, University of Arizona, 933 N. Cherry Ave., Tucson, AZ, 85721, USA}
\altaffiltext{17}{Physics Department, Faculty of Science, Hokkaido University, Kita-10 Nishi 8, Kita-ku, Sapporo 060-0810, Japan}
\altaffiltext{18}{ETH Zurich Institute for Astronomy. CH-8093 Zurich. Switzerland}
\altaffiltext{19}{Department of Physics, Stanford University, Stanford, CA 94305, USA}
\altaffiltext{20}{Department of Astrophysical Sciences, Princeton University, Peyton Hall, Princeton, NJ 08544, USA}

>>>>

\title{Enzo: An Adaptive Mesh Refinement Code for Astrophysics}
\author{The Enzo Collaboration:}
\author{Greg L. Bryan\altaffilmark{1}}
\author{Michael L. Norman\altaffilmark{2,3}}
\author{Brian W. O'Shea\altaffilmark{4,5}}
\author{Tom Abel\altaffilmark{6, 20}}
\author{John H. Wise\altaffilmark{7}}
\author{Matthew J. Turk\altaffilmark{1}}
\author{Daniel R. Reynolds\altaffilmark{8}}
\author{David C. Collins\altaffilmark{9}}
\author{Peng Wang\altaffilmark{6}}
\author{Samuel W. Skillman\altaffilmark{10,11}}
\author{Britton Smith\altaffilmark{4}}
\author{Robert P. Harkness\altaffilmark{12}}
\author{James Bordner\altaffilmark{2}}
\author{Ji-hoon Kim\altaffilmark{13}}
\author{Michael Kuhlen\altaffilmark{14,15}}
\author{Hao Xu\altaffilmark{2}}
\author{Nathan Goldbaum\altaffilmark{15}}
\author{Cameron Hummels\altaffilmark{16}}
\author{Alexei G. Kritsuk\altaffilmark{2}}
\author{Elizabeth Tasker\altaffilmark{17}}
\author{Stephen Skory\altaffilmark{2}}
\author{Christine Simpson\altaffilmark{1}}
\author{Oliver Hahn\altaffilmark{18}}
\author{Jeffrey S. Oishi\altaffilmark{19}}
\author{Geoffrey C So\altaffilmark{2}}
\author{Fen Zhao\altaffilmark{20}}
\author{Renyue Cen\altaffilmark{21}}
\author{Yuan Li\altaffilmark{1}}

\altaffiltext{1}{Columbia University, Department of Astronomy, New York, NY, 10025, USA}
\altaffiltext{2}{CASS, University of California, San Diego, 9500 Gilman Drive La Jolla, CA 92093-0424, USA}
\altaffiltext{3}{SDSC, University of California, San Diego, 9500 Gilman Drive La Jolla, CA 92093-0505, USA}
\altaffiltext{4}{Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\altaffiltext{5}{Lyman Briggs College and Institute for Cyber-Enabled Research, Michigan State University, East Lansing, MI 48824, USA}
\altaffiltext{6}{Kavli Institute for Particle Astrophysics and Cosmology, Stanford University, Menlo Park, CA 94025, USA}
\altaffiltext{7}{Center for Relativistic Astrophysics, School of Physics, Georgia Institute of Technology, 837 State Street, Atlanta, GA, USA}
\altaffiltext{8}{Southern Methodist University, 6425 Boaz Ln, Dallas, TX 75205, USA}
\altaffiltext{9}{Theoretical Division, Los Alamos National Laboratory, Los Alamos, NM 87544}
\altaffiltext{10}{Center for Astrophysics and Space Astronomy, Department of Astrophysical and Planetary Science, University of Colorado, Boulder, CO 80309, USA}
\altaffiltext{11}{DOE Computational Science Graduate Fellow}
\altaffiltext{12}{NICS, Oak Ridge National Laboratory, 1 Bethel Valley Rd, Oak Ridge, TN 37831, USA}
\altaffiltext{13}{Department of Astronomy and Astrophysics, University of California, Santa Cruz, CA 95064, USA}
\altaffiltext{14}{Theoretical Astrophysics Center, University of California Berkeley, Hearst Field Annex, Berkeley, CA 94720, USA}
\altaffiltext{15}{Department of Astronomy and Astrophysics, University of California, 1156 High Street, Santa Cruz, CA 95064, USA}
\altaffiltext{16}{Department of Astronomy/Steward Observatory, University of Arizona, 933 N. Cherry Ave., Tucson, AZ, 85721, USA}
\altaffiltext{17}{Physics Department, Faculty of Science, Hokkaido University, Kita-10 Nishi 8, Kita-ku, Sapporo 060-0810, Japan}
\altaffiltext{18}{ETH Zurich Institute for Astronomy. CH-8093 Zurich. Switzerland}
\altaffiltext{19}{Department of Astrophysics, American Museum of Natural History, Central Park West at 81st St, New York, NY 10024}
\altaffiltext{20}{Department of Physics, Stanford University, Stanford, CA 94305, USA}
\altaffiltext{21}{Department of Astrophysical Sciences, Princeton University, Peyton Hall, Princeton, NJ 08544, USA}

### 2013-06-17 numerical-analysis.tex
Jeff Oishi's changes.

We expose \enzo's mesh geometry and fluid quantities to the analysis platform
\texttt{yt} \citep{2011ApJS..192....9T, 2011arXiv1112.4482T}.  At compile time,
\enzo\ is (dynamically or statically) linked against the Python and NumPy
libraries necessary to create proxy objects exposing the mesh geometry,
fluid quantities and particle arrays as NumPy arrays.  This information is then
passed to a special handler inside \texttt{yt}.  \texttt{yt} interprets the
mesh and fluid information and, without saving data to disk except as
explicitly requested, constructs a native representation of the in-memory state
of the simulation that appears identical to an on-disk simulation output.
\enzo\ then executes a user-provided analysis script, which is able to access the
in-memory simulation object.  Once the analysis script has returned control to
\enzo, the simulation proceeds.  This process can occur at either the top of the
main ``EvolveHierarchy'' loop or at the end of a timestep at the finest level,
and the frequency with which it is called is adjustable by a run-time
parameter.  During the course of conducting analysis, the simulation is halted
until the conclusion of the analysis.

@@

\begin{eqnarray}
\nabla \cdot \vec{v} < 0,\\
\nabla T \cdot \nabla S > 0,\\
T_2 > T_1,\\
\rho_2 > \rho_1,
\end{eqnarray}

where $\vec{v}$ is the velocity field, $T$ is the temperature, $\rho$ is the
density, and $S=T/\rho^{\gamma-1}$ is the entropy.  $T_2$ and $T_1$ are the post-shock~(downstream)
and pre-shock~(upstream) temperatures, respectively. An optional temperature floor
may be chosen, which is useful for situations such as cosmological simulations without a radiation
background where underdense gas in the intergalactic medium (IGM) can cool adiabatically to
very low temperatures.

>>>>

We expose \enzo's mesh geometry and fluid quantities to the analysis platform
\texttt{yt} \citep{2011ApJS..192....9T, 2011arXiv1112.4482T}.  At compile time,
\enzo\ is (dynamically or statically) linked against the Python and NumPy
libraries necessary to create proxy objects exposing the mesh geometry,
fluid quantities and particle arrays as NumPy arrays.  This information is then
passed to a special handler inside \texttt{yt}.  \texttt{yt} interprets the
mesh and fluid information and, without saving data to disk, constructs a native representation of the in-memory state
of the simulation that appears identical to an on-disk simulation output.
\enzo\ then executes a user-provided analysis script, which is able to access the
in-memory simulation object.  Once the analysis script has returned control to
\enzo, the simulation proceeds.  This process can occur at either the top of the
main ``EvolveHierarchy'' loop or at the end of a timestep at the finest level,
and the frequency with which it is called is adjustable by a run-time
parameter.  During the course of conducting analysis, the simulation is halted
until the conclusion of the analysis.

@@

\begin{eqnarray}
\div \myvec{v} < 0,\\
\grad T \cdot \grad S > 0,\\
T_2 > T_1,\\
\rho_2 > \rho_1,
\end{eqnarray}

where $\myvec{v}$ is the velocity field, $T$ is the temperature, $\rho$ is the
density, and $S=T/\rho^{\gamma-1}$ is the entropy.  $T_2$ and $T_1$ are the post-shock~(downstream)
and pre-shock~(upstream) temperatures, respectively. An optional temperature floor
may be chosen, which is useful for situations such as cosmological simulations without a radiation
background where underdense gas in the intergalactic medium (IGM) can cool adiabatically to
very low temperatures.

### 2013-06-17 numerical-overview.tex
Jeff Oishi's changes.

The system of equations described above works well for systems with
relatively low Mach numbers, as long as these systems are well resolved.  
However, cosmology is replete with situations where there are bulk hypersonic
flows.  In these situations, the ratio of kinetic to thermal energy can be
very high -- in some situations up to $10^6 - 10^8$.  This implies that the
thermal energy is an extremely tiny fraction of the kinetic energy, which can 
cause numerical problems when one is interested in just the thermal energy of the gas, since
 Equation~\ref{enzoenergy} solves for the total energy.  In this system of
equations, the
 thermal energy $E_{therm}$ is calculated as $E - E_{kin}$, 
where E is the total specific energy as
calculated in equation~\ref{enzoenergy} and $E_{kin}$ is the specific kinetic energy,
$0.5 \vec{v}_b^2$.  
In hypersonic flows $E$ and $E_{kin}$ are nearly the same, and any 
number calculated as 
the difference of these is going to be strongly affected by numerical error.  To 
avoid this problem, \enzo\ also solves the internal energy equation in
comoving coordinates:

\begin{equation}
\frac{\partial e}{\partial t} + \frac{1}{a} \vec{v}_b \cdot \grad e = 
- \frac{3(\gamma-1)\dot{a}}{a}e - \frac{p}{a \rho} \div \vec{v}_b
\label{enzointenergy}
\end{equation}

@@

\begin{equation}
\div \vec{v}_b < 0,
\label{cencont}
\end{equation}

@@

where $\vec{v}$ is the velocity of the gas in the cell, $\rho_{b}$ and 
$\rho_{d}$ are the cell's baryon and dark matter density, respectively,
$\rho_{total} = \rho_{b} + \rho_{d}$, $m_{b}$ and $m_{j}$ are the 
baryonic mass in the cell and jeans mass of the cell, and C is the 
isothermal sound speed in the cell.  
If all of these criteria are met, 
the mass of a star particle is calculated as \(m_{*} = m_{b} 
\frac{ \Delta t}{ t_{dyn} } f_{*eff} \), 
where $f_{*eff}$ is the star formation efficiency parameter.

@@

\begin{equation}
\Delta t_{accel} = min \left( \sqrt{\frac{\Delta x}{\vec{g}}} \right)_L 
\label{eqn:dtaccel}
\end{equation}

@@

Equation~\ref{eqn:dtaccel} is supplementary to equation~\ref{eqn:dthydro} in that it
takes into account the possibility of large accelerations causing numerical 
instabilities by violating the Courant condition.  In this equation, $\vec{g}$ is the
gravitational acceleration in each cell on level L.  \blue{ \emph{It should be
  noted that the timestep is computed before the acceleration field is computed,
  and the acceleration field is deleted at the end of each timestep.
  So this condition isn't ever actually enforced.  We should either
  fix this problem, or not discuss it.
 -dcollins.} }

>>>>

The system of equations described above works well for systems with
relatively low Mach numbers, as long as these systems are well resolved.  
However, cosmology is replete with situations where there are bulk hypersonic
flows.  In these situations, the ratio of kinetic to thermal energy can be
very high -- in some situations up to $10^6 - 10^8$.  This implies that the
thermal energy is an extremely tiny fraction of the kinetic energy, which can 
cause numerical problems when one is interested in just the thermal energy of the gas, since
 Equation~\ref{enzoenergy} solves for the total energy.  In this system of
equations, the
 thermal energy $E_{therm}$ is calculated as $E - E_{kin}$, 
where E is the total specific energy as
calculated in equation~\ref{enzoenergy} and $E_{kin}$ is the specific kinetic energy,
$0.5 \myvec{v}_b^2$.  
In hypersonic flows $E$ and $E_{kin}$ are nearly the same, and any 
number calculated as 
the difference of these is going to be strongly affected by numerical error.  To 
avoid this problem, \enzo\ also solves the internal energy equation in
comoving coordinates:

\begin{equation}
\frac{\partial e}{\partial t} + \frac{1}{a} \myvec{v}_b \cdot \grad e = 
- \frac{3(\gamma-1)\dot{a}}{a}e - \frac{p}{a \rho} \div \myvec{v}_b
\label{enzointenergy}
\end{equation}

@@

\begin{equation}
\div \myvec{v}_b < 0,
\label{cencont}
\end{equation}

@@

where $\myvec{v}$ is the velocity of the gas in the cell, $\rho_{b}$ and 
$\rho_{d}$ are the cell's baryon and dark matter density, respectively,
$\rho_{total} = \rho_{b} + \rho_{d}$, $m_{b}$ and $m_{j}$ are the 
baryonic mass in the cell and jeans mass of the cell, and C is the 
isothermal sound speed in the cell.  
If all of these criteria are met, 
the mass of a star particle is calculated as \(m_{*} = m_{b} 
\frac{ \Delta t}{ t_{dyn} } f_{*eff} \), 
where $f_{*eff}$ is the star formation efficiency parameter.

@@

\begin{equation}
\Delta t_{accel} = min \left( \sqrt{\frac{\Delta x}{\myvec{g}}} \right)_L 
\label{eqn:dtaccel}
\end{equation}

@@

Equation~\ref{eqn:dtaccel} is supplementary to equation~\ref{eqn:dthydro} in that it
takes into account the possibility of large accelerations causing numerical 
instabilities by violating the Courant condition.  In this equation, $\myvec{g}$ is the
gravitational acceleration in each cell on level L.  \blue{ \emph{It should be
  noted that the timestep is computed before the acceleration field is computed,
  and the acceleration field is deleted at the end of each timestep.
  So this condition isn't ever actually enforced.  We should either
  fix this problem, or not discuss it.
 -dcollins.} }

### 2013-06-17 numerical-timestep.tex
A few more edites to section 2. Also unified "time step" -> "timestep" throughout the whole manuscript.
Jeff Oishi's changes.

\section{Time Stepping}
\label{sec.timestepping}

@@

\begin{equation}
\Delta t_{\rm accel} = \min \left( \sqrt{\frac{\Delta x}{\vec{g}}} \right)_L ,
\label{eqn:dtaccel}
\end{equation}

\begin{equation}
\Delta t_{\rm rad} = \min \left(  \sqrt{\frac{\Delta x}{\vec{a}_{rad}}} \right)_L,
\label{eqn:dtrad}
\end{equation}

@@

Equations~(\ref{eqn:dtaccel}) and~(\ref{eqn:dtrad}) are supplementary to equation~(\ref{eqn:dthydro}) in that they
take into account the possibility of large accelerations due to either
gravity (equation~\ref{eqn:dtaccel}) or radiation pressure
(equation~\ref{eqn:dtrad}).  In equation~(\ref{eqn:dtaccel}), $\vec{g}$ is the
gravitational acceleration in each cell on level l.  In
equation~(\ref{eqn:dtrad}), $\vec{a}_{\rm rad}$ is the estimated
acceleration due to radiation pressure in each cell on level l,
defined as
\begin{equation}
\vec{a}_{rad} = \frac{ \sum_i \frac{\dot{E_i}}{c} \hat{r_i} }{m_b} 
\end{equation}
where the sum calculates the energy deposited in a cell during the
previous timestep due to \textit{all photon packets} that crossed that
cell, with $\hat{r_i}$ being a unit vector that accounts for the
packet direction.

>>>>

\section{Timestepping}
\label{sec.timestepping}

@@

\begin{equation}
\Delta t_{\rm accel} = \min \left( \sqrt{\frac{\Delta x}{\myvec{g}}} \right)_L ,
\label{eqn:dtaccel}
\end{equation}

\begin{equation}
\Delta t_{\rm rad} = \min \left(  \sqrt{\frac{\Delta x}{\myvec{a}_{rad}}} \right)_L,
\label{eqn:dtrad}
\end{equation}

@@

Equations~(\ref{eqn:dtaccel}) and~(\ref{eqn:dtrad}) are supplementary to equation~(\ref{eqn:dthydro}) in that they
take into account the possibility of large accelerations due to either
gravity (equation~\ref{eqn:dtaccel}) or radiation pressure
(equation~\ref{eqn:dtrad}).  In equation~(\ref{eqn:dtaccel}), $\myvec{g}$ is the
gravitational acceleration in each cell on level l.  In
equation~(\ref{eqn:dtrad}), $\myvec{a}_{\rm rad}$ is the estimated
acceleration due to radiation pressure in each cell on level l,
defined as
\begin{equation}
\myvec{a}_{rad} = \frac{ \sum_i \frac{\dot{E_i}}{c} \hat{r_i} }{m_b} 
\end{equation}
where the sum calculates the energy deposited in a cell during the
previous timestep due to \textit{all photon packets} that crossed that
cell, with $\hat{r_i}$ being a unit vector that accounts for the
packet direction.
